---
title: "Advanced statistics for physics analysis - Exercise 6"
author: "Matteo Bortoletto, matr. 1242935"
date: "15/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1
The number of particles emitted by a radioactive source during a fixed interval of time ($\Delta t = 10$ s) follows a Poisson distribution on the parameter $\mu$. The number of particles observed during consecutive time intervals is: 4, 1, 3, 1 and 3.

a. Suppose a uniform prior distribution for the parameter $\mu$.

    - Determine and draw the posterior distribution for $\mu$, given the data. 
    - Evaluate mean, median and variance, both analytically and numerically in `R`.

```{r}
#code
```

b. Suppose a Jeffrey’s prior for the parameter $\mu$. 

    - Determine and draw the posterior distribution for $\mu$, given the data. 
    - Evaluate mean, median and variance, both analytically and numerically in `R`.

```{r}
#code
```

c. Evaluate a 95% credibility interval for the results obtained with both priors. Compare the result with that obtained using a normal approximation for the posterior distribution, with the same mean and standard deviation.

```{r}
#code
```


# Exercise 2
Given the problem of the lighthouse discussed last week, study the case in which both the position along the shore ($\alpha$) and the distance out at the sea ($\beta$) are unknown.

```{r}
#code
```


# Exercise 3
Given the Signal over Background example discussed last week, analyze and discuss the following cases:

a. Vary the sampling resolution used to generate the data, keeping the same sampling range:

`xdat <- seq(from=-7*w, to=7*w, by=0.5*w)`

Use $w = \{0.1,0.25,1,2,3\}$. Check the effect on the results.

```{r}
#code
```

b. Change the ratio $A/B$ used to simulate the data (keeping both positive in accordance with the prior). Check the effect on the results.

```{r}
#code
```
