---
title: "Advanced statistics for physics analysis - Exercise 7"
author: "Matteo Bortoletto, matr. 1242935"
date: "30/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1
A researcher has collected $n = 15$ observations that are supposed to come from a Normal distribution with known variance $\sigma^2 = 16$:

`26.8, 26.3, 28.3, 28.5, 16.3, 31.9, 28.5, 27.2, 20.9, 27.5, 28.0, 18.6, 22.3, 25.0, 31.5`

Assuming a normal prior for $\mu$, $\mathcal{N}(m=20, s^2=25)$:

a. Determine the posterior distribution $P(\mu | y_1,\dots,y_{15})$ and find the posterior mean and standard deviation.

```{r}
library(latex2exp)

s.known <- 4
N <- 15
data <- c(26.8, 26.3, 28.3, 28.5, 16.3, 31.9, 28.5, 27.2, 20.9, 27.5, 28.0, 18.6, 22.3, 25.0, 31.5)
m.data <- mean(data)
#cat("Mean of the data = ", m.data, "\n")
s.prior <- 5
m.prior <- 20
s.post <- s.known*s.prior/(sqrt(s.known^2+N*s.prior^2))
#m.post <- (1/s.prior^2)*m.prior/(1/s.known^2 + 1/s.prior^2) + (N/s.known^2)*m.data/(N/s.known^2 + 1/s.prior^2)
#s.post <- sqrt((1/s.prior))
m.post <- (s.post^2)*(m.prior/s.prior^2 + m.data*N/s.known^2)

mu <- seq(1, 60, 0.1)
post <- dnorm(mu, m.post, s.post)

cat("Posterior mean =", m.post, "\n")
cat("Posterior standard deviation =", s.post)
```
The mean of the posterior is $\bar{\mu} = `r m.post`$ and the standard deviation is $\sigma_{post} = `r s.post`$.

b. Find the 95% credibility interval for $\mu$.

```{r}
mu1 <- mu[which.min(abs(0.025-cumsum(post)*0.1))]
mu2 <- mu[which.min(abs(0.975-cumsum(post)*0.1))]
cat("The 95% credibility interval is given by [1, x2] = [", mu1, ",", mu2, "]")
```
The 95% credibility interval is given by $[\mu_1,\mu_2] = [`r mu1`, `r mu2`]$.

c. Plot the posterior distribution, indicating on the same plot: the mean value, the standard deviation, and the 95% credibility interval.

```{r}
options(repr.plot.width=12, repr.plot.height=10)

mu.p <- seq(mu1, mu2, 0.01)
x <- c(mu1, mu.p, mu2)
post.95 <- dnorm(mu.p, m.post, s.post)
y <- c(0, post.95, 0)

plot(mu, post,
     type = "l",
     xlim = c(20, 31),
     col = "royalblue4",
     xlab = TeX("$\\mu$"),
     ylab = TeX("P($\\mu$ | y_1,...,y_{15})"),
     main = "Posterior distribution \n")

text(30, 0.38, paste("mean =", round(m.post,0)))
text(30, 0.34, paste("std =", round(s.post,1)))
abline(v = m.post, lty = 2, col = "royalblue3")
polygon(x=x, y=y, col='royalblue', pch=1, density = 10)

abline(v=m.post, col="royalblue", lty=2, pch=1)
abline(v=mu1, col="royalblue", lty=2, pch=1)
abline(v=mu2, col="royalblue", lty=2, pch=1)
text(m.post, par("usr")[4] + 0.035, labels = TeX("$\\bar{\\mu}$"), pos = 1, cex = 1.2, col="black", xpd=TRUE)
text(mu1, par("usr")[4] + 0.035, labels = TeX("$\\mu_{2.5%}$"), pos = 1, cex = 1.2, col="black", xpd=TRUE)
text(mu2, par("usr")[4] + 0.035, labels = TeX("$\\mu_{97.5%}$"), pos = 1, cex = 1.2, col="black", xpd=TRUE)
```

d. repeat the analysis using a different prior $\mathcal{N}(m=30, s^2=16)$ and plot, on the same graph the likelihood, the prior and the posterior.

```{r}
s.prior <- 4
m.prior <- 30

prior.new <- dnorm(mu, m.prior, s.prior)

likelihood <- Vectorize(function(mu, s) {
  l <- 1/(sqrt(2*pi)*s)*exp(-(sum(data-mu)^2)/(2*s^2))
  return(l)
})

s.post <- s.known*s.prior/(sqrt(s.known^2+N*s.prior^2))
#m.post <- (1/s.prior^2)*m.prior/(1/s.known^2 + 1/s.prior^2) + (N/s.known^2)*m.data/(N/s.known^2 + 1/s.prior^2)
m.post <- (s.post^2)*(m.prior/s.prior^2 + m.data*N/s.known^2)

likelihood.norm <- likelihood(mu, s.known)/(sum(likelihood(mu, s.known))*0.1)

post.new <- dnorm(mu, m.post, s.post)/sum(dnorm(mu, m.post, s.post)*0.1)

#post.new1 <- prior.new*likelihood(x, s.known)/(sum(prior.new*likelihood(x, s.known))*0.1)

plot(mu, likelihood.norm,
     xlim = c(20,40),
     type = "l",
     lwd = 1.5,
     col = "royalblue",
     xlab = TeX("$\\mu$"),
     ylab = "Posterior",
     main = "Posterior distribution")

lines(mu, prior.new,
      lwd = 1.5,
      type = "l",
      col = "coral2")

lines(mu, post.new,
      lwd = 1.5,
      type = "l",
      col = "palegreen4")

legend("topright", c("Likelihood", "Prior", "Posterior"), col = c("royalblue", "coral2", "palegreen4"), lwd = 3)
```

e. Compare the credibility intervals obtained with the two priors.

```{r}
mu1.new <- mu[which.min(abs(0.025-cumsum(post.new)*0.1))]
mu2.new <- mu[which.min(abs(0.975-cumsum(post.new)*0.1))]
cat("The 95% credibility interval is given by [x1, x2] = [", mu1.new, ",", mu2.new, "]")
```
The 95% credibility interval is given by $[\mu_1,\mu_2] = [`r mu1.new`, `r mu2.new`]$.


## Exercise 2
A researcher has collected $n = 16$ observations that are supposed to come from a Normal distribution with known variance $\sigma^2 = 4$:

`4.09, 4.68, 1.87, 2.62, 5.58, 8.68, 4.07, 4.78, 4.79, 4.49, 5.85, 5.09, 2.40, 6.27, 6.30, 4.47`

Assuming the prior is a step function:

$$
g(\mu) = 
\begin{cases}
  \mu   & \text{for } 0<\mu\le 3 \\
  3     & \text{for } 3<\mu\le 5 \\
  8-\mu & \text{for } 5<\mu\le 8 \\
  0     & \text{for } \mu > 8
\end{cases}
$$

a. Find the posterior distribution, the posterior mean and standard deviation.

```{r}
n <- 16
s.known <- 2
data <- c(4.09, 4.68, 1.87, 2.62, 5.58, 8.68, 4.07, 4.78, 4.79, 4.49, 5.85, 5.09, 2.40, 6.27, 6.30, 4.47)

prior <- function(mu) {
  val <- ifelse(0 < mu & mu <= 3, mu,
                ifelse(3 < mu & mu <= 5, 3,
                       ifelse(5 < mu & mu <= 8, 8 - mu, 0)))
  return(val)
}

mu <- seq(0,10,0.01)

#like <- dnorm(mu, mean = mean(data), sd = s.known)/sum(dnorm(mu, mean = mean(data), sd = s.known)*0.01)

likelihood <- Vectorize(function(mu, s) {
  l <- 1/(sqrt(2*pi)*s)*exp(-(sum(data-mu)^2)/(2*s^2))
  return(l)
})

likelihood.norm <- likelihood(mu, s.known)/(sum(likelihood(mu, s.known))*0.01)
post.star <- likelihood.norm*prior(mu)
post.norm <- post.star/(sum(post.star)*0.01)

mu.max <- mu[which.max(post.norm)]

cat("The maximum of the posterior is:", mu.max)
```
The maximum of the posterior is $\mu_{max} = `r mu.max`$.

b. Find the 95% credibility interval for $\mu$.

```{r}
mu1 <- mu[which.min(abs(0.025-cumsum(post.norm)*0.01))]
mu2 <- mu[which.min(abs(0.975-cumsum(post.norm)*0.01))]
cat("The 95% credibility interval is given by [mu1, mu2] = [", mu1, ",", mu2, "]")
```
The 95% credibility interval is given by $[\mu_1,\mu_2] = [`r mu1`, `r mu2`]$.

c. Plot the posterior distribution, indicating on the same plot: the mean value, the standard deviation, and the 95% credibility interval.

```{r}
library(latex2exp)
options(repr.plot.width=12, repr.plot.height=10)

mu.p <- seq(mu1, mu2, 0.01)
x <- c(mu1, mu.p, mu2)
likelihood.norm.95 <- likelihood(mu.p, s.known)/(sum(likelihood(mu.p, s.known))*0.01)
post.star.95 <- likelihood.norm.95*prior(mu.p)
post.norm.95 <- post.star.95/(sum(post.star.95)*0.01)
y <- c(0, post.norm.95, 0)

plot(mu, post.norm, 
     type='l', 
     lwd=1.5, 
     xlim = c(4,5.5),
     ylim = c(0,4),
     col = "royalblue4",
     xlab = TeX("$\\mu$"),
     ylab = TeX("P($\\mu$ | y_1,...,y_{16})"),
     main='Posterior \n',
     pch=2)

polygon(x=x, y=y, col='royalblue', pch=1, density = 10)
abline(v=mu.max,    col="grey40", lty=2, pch=1)
abline(v=mu1, col="grey40", lty=2, pch=1)
abline(v=mu2, col="grey40", lty=2, pch=1)
text(mu.max, par("usr")[4] + 0.4, labels = TeX("$\\bar{\\mu}$"), pos = 1, cex = 1.2, col="black", xpd=TRUE)
text(mu1, par("usr")[4] + 0.4, labels = TeX("$\\mu_{2.5%}$"), pos = 1, cex = 1.2, col="black", xpd=TRUE)
text(mu2, par("usr")[4] + 0.4, labels = TeX("$\\mu_{97.5%}$"), pos = 1, cex = 1.2, col="black", xpd=TRUE)

#text(5.4, 4.8, paste("mean =", round(m.post,0)))
#text(5.4, 4.4, paste("std =", round(s.post,1)))
```

d. Plot, on the same graph, the prior, the likelihood and the posterior distribution.

```{r}
plot(mu, post.norm,
     type = "l",
     lty = 2,
     lwd = 1.5,
     col = "blue",
     xlab = TeX("$\\mu$"),
     ylab = "P",
     main = "Posterior, prior and likelihood distributions")

lines(mu, prior(mu)/15, # normalization
      type = "l",
      lwd = 1.5,
      col = "red")

lines(mu, likelihood.norm,
      type = "l",
      lwd = 1.5,
      col = "darkolivegreen4")

legend("topright", 
       c("Posterior", "Prior", "Likelihood"), 
       lwd = 3, 
       col = c("blue", "red", "darkolivegreen4"))
```


## Exercise 3
In a study on water quality of streams, a high level of bacter X was defined as a level greater than 100 per 100 ml of stream water. $n = 116$ samples were taken from streams having a high environmental impact on pandas. Out of these, $y = 11$ had a high bacter X level.

Indicating with $p$ the probability that a sample of water taken from the stream has a high bacter X level:

a. Find the frequentist estimator for $p$.

```{r}
n <- 116
y <- 11
p.freq <- 11/116
cat("The frequentist estimator is p =", p.freq)
```
The frequentist estimator for $p$ is $\hat{p} = `r p.freq`$.

b. Using a Beta$(1,10)$ prior for $p$, calculate and plot posterior distribution $P(p|y)$.

```{r}
alpha <- 1
beta <- 10
n.sample <- 2000
delta <- 1/n.sample
p <- seq(from=1/(2*n.sample), by=1/n.sample, length.out=n.sample)

prior <- dbeta(p, alpha, beta)

likelihood <- dbinom(y, n, p)

post.star <- dbeta(p, alpha + y, beta + n - y)
post.norm <- post.star/(sum(post.star)*delta) 

plot(p, post.norm, 
     type = "l",
     lwd = 1.5,
     col = "red",
     ylab = "P(p|y)",
     xlab = "p",
     main = "Posterior")
```

c. Find the bayesian estimator for $p$, the posterior mean and variance, and a 95% credible interval.

```{r}
post.mean <- (alpha + y)/(alpha + beta + n)
#post.m <- sum(p*post.norm)/n.sample
post.var <- (1/(n+2)^2)*n*post.mean*(1-post.mean)
cat("The bayesian estimator is the posterior mean:", post.mean, "\n")
cat("The variance is:", post.var)
```
The bayesian estimator is the posterior mean, $\bar{p} = `r post.mean` and the variance is Var $ = `r post.var`. 

d. test the hypothesis $H_0 : p = 0.1$ versus $H_1 : p \ne 0.1$ at 5% level of significance with both the frequentist and bayesian approaches.

```{r}

```

A new measurement, performed one month later on $n = 165$ water samples, gives $y = 9$ high bacter X level.

e. Find the frequentist estimator for $p$.

```{r}

```

f. Find a bayesian estimator for $p$, assuming both a Beta$(1,10)$ prior for $p$, and the posterior probability of the older measurement as the prior for the new one.

```{r}

```

g. Find the bayesian estimator for $p$, the posterior mean and variance, and a 95% credible interval.

```{r}

```

h. test the hypothesis $H_0 : p = 0.1$ versus $H_1 : p \ne 0.1$ at 5% level of significance with both the frequentist and bayesian approaches.

```{r}

```