---
title: "Advanced statistics for physics analysis - Exercise 4"
author: "Matteo Bortoletto, matr. 1242935"
date: "28/04/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 1
The triangular distribution in the interval $(a, b)$, is given by the following:

$$
f(x) = \begin{cases}
  \frac{2(x-a)}{(b-a)(c-a)} & a\le x < c \\
  \frac{2(b-x)}{(b-a)(b-c)} & c\le x \le b \\
  0 & \text{otherwise}
\end{cases}
$$
where $c \in [a, b]$.

a. Plot the function, given the interval (a, b);
```{r}
f <- function(x, a, b, c) {
  func <- ifelse(a<=x & x<c, 2*(x-a)/((b-a)*(c-a)), 
                 ifelse(c<=x & x<=b, 2*(b-x)/((b-a)*(b-c)), 0))
  return(func)
}

x <- seq(-5,5,0.001)
a <- -2
b <- 3
c <- 1

plot(x, f(x,a,b,c),
     type = "l",
     xlab = "x",
     ylab = "f(x)",
     main = "Triangular distribution",
     col = "forestgreen")
```
b. Write an algorithm to generate random numbers from the triangular distribution;
```{r}
set.seed(5)

cdf_f <- function(x,a,b,c) {
  func <- ifelse(x<a, 0,
                 ifelse(x<=c, (a-x)^2/((a-b)*(a-c)), 
                        ifelse(x<=b, (x-a)*(a-2*b+x)/((a-b)*(b-c)), 1)))
  return(func)
}

q_f <- function(x,a,b,c) {
  func <- ifelse(x<=cdf_f(c,a,b,c), a+sqrt((b-a)*(c-a)*x),
               ifelse(x>cdf_f(c,a,b,c), b-sqrt((1-x)*(b-a)*(b-c)), 0))
  return(func)
}

r_f <- function(n,a,b,c) {
    func <- q_f(runif(n),a,b,c)
    return(func)
}
```

c. Generate $10^4$ random numbers from the distribution, show them in an histogram and superimpose the analytical curve.
```{r}
set.seed(5)

a <- -3
b <- 2
c <- 1
f_samples <- r_f(10000,a,b,c)
f_samples[1:10]

hist(f_samples, 
     xlab = "x",
     ylab = "f(x)",
     main = "Triangular distribution histogram",
     col = "aquamarine3",
     density = 50,
     freq = F,
     ylim = c(0,0.45))

lines(seq(a,b,0.001), f(seq(a,b,0.001),a,b,c),
      col = "forestgreen")

legend("topright", c("f(x)"), lwd=4, col="forestgreen")
```

# Exercise 2
Given a discrete probability distribution defined by the following probabilities: 0.05, 0.19, 0.14, 0.17, 0.02, 0.11, 0.06, 0.05, 0.04, 0.17:

a. Plot the probability density function and the cumulative density function;
```{r}
p <- c(0.05, 0.19, 0.14, 0.17, 0.02, 0.11, 0.06, 0.05, 0.04, 0.17)
csum <- cumsum(p)

plot(seq(1,10), p,
     xlab = "x",
     ylab = "Probability",
     type = "s",
     col = "forestgreen",
     main = "PDF")

plot(seq(1,10), csum,
     xlab = "x",
     ylab = "CDF",
     type = "s",
     col = "forestgreen",
     main = "CDF")
```
b. write an algorithm to generate random numbers from the discrete probability distribution
```{r}
sampl <- sample(x=seq(1,10), size=10000, replace=TRUE, prob=p)
sampl[1:10]
#####################################
#####################################
#####################################
####### TROVA UN ALTRO MODO #########
#####################################
#####################################
#####################################
```

# Exercise 3
Generate random variables from the following distribution:

$$
f(x) = \frac{2}{\pi R^2}\sqrt{R^2-x^2}
$$
where $−R\le x \le R$. Using the acceptance-rejection algorithm, assume $M = 2/(\pi R)$ and generate $10^4$ random variables, plotting them in an histogram.
```{r}

```

# Exercise 4
An important property of the gamma distribution is the so-called reproductive property: given a sequence of independent random variables $X_j \sim \Gamma(\alpha_j,\beta)$, it follows that

$$
Y = \sum_{j=1}^n X_j \longrightarrow \Gamma(\alpha,\beta), \quad \text{where } \alpha = \sum_{j=1}^n \alpha_j
$$
If $\alpha = m$ is an integer, a random variable from gamma distribution $\Gamma(m,\beta)$ (also known as Erlang distribution) can be obtained by summing $m$ independent exponential random variables $X_j \sim \text{Exp}(\beta)$:

$$
Y = \beta\sum_{j=1}^n (-\ln U_j) = -\beta\prod_{j=1}^n U_j
$$
write an algorithm to sample variables from an Erlang distribution $\Gamma(m,\beta)$.
```{r}

```

# Exercise 5
One of the first random number generators was proposed by von Neumann, the so-called middle square algorithm. Write R code to implement this type of generator and, given a fixed digit number input, square it an remove the leading and trailing digits, in order to return a number with the same number of digits as the original number. 

*Suggestion*: after having squared the number, convert it to a list of characters 

`(number <- unlist(strsplit(as.character(x.squared),"")))`

and, after having removed the head and tail of the list, convert it back to a number

`(as.numeric(paste(number.after.trimming, collapse="")))`
```{r}

```